---
title: "Testing My LDA"
author: "Georgie Mansell"
date: "02/08/2021"
output: pdf_document
geometry: margin=0.8in
urlcolor: purple
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=T, fig.align="center", out.width="80%")
knitr::opts_knit$set(root.dir = '/home/an20830/Documents/COMPASS/Research Project/3_MyLDAPackage/MSLDA')
```

# Introduction
MSLDA is now an R package on github (https://github.com/g-l-mansell/MSLDA) which contains my different implementations of LDA:
```{r, message=F}
library(MSLDA)
library(tidyverse)
```

* lda_original - as in the Blei 2003 paper, but edited to force $\alpha_i > 0$ - the only implementation that does not run in parallel

* lda_original_par - as above but the E-step is run in parallel - should give the same results with the same seed.

* lda_noalpha - since the alpha update rule in LDA_original is flawed, this version treats alpha as a hyperparameter which should be tuned.

* lda_reshaped - this version is a further adaptation of LDA_original, which uses a count (document-term) matrix as an input rather than the document vectors - this should give the same results as LDA_originial_par with improved speed.

* lda_reshaped_noalpha - this is the reshaped algorithm above but with alpha as a hyperparamter - this should give the same results as LDA_noalpha with improved speed.

* lda_smoothed - as in the Hoffman 2010 paper (batch LDA section), with edited equation for $\mathcal{L}$ - this version assumes beta is a random variable.

# Check all 5 implementations work
using the simulated dataset of documents and the same seed.

```{r, fig.width=8, fig.height=3, message=F}
load("data/MyCorpus.Rdata")
par(mfrow=c(1, 2))

res1 <- lda_original(docs, K=3, seed=83)
plot(res1$Ls, ylab="L", main="Original")

res2 <- lda_original_par(docs, K=3, seed=83)
plot(res2$Ls, ylab="L", main="Original Parallel")

res3 <- lda_noalpha(docs, K=3, seed=83)
plot(res3$Ls, ylab="L", main="Original no alpha")

res4 <- lda_reshaped(counts, K=3, seed=83)
plot(res4$Ls, ylab="L", main="Reshaped")

res5 <- lda_reshaped_noalpha(counts, K=3, seed=83)
plot(res4$Ls, ylab="L", main="Reshaped no alpha")

res6 <- lda_smoothed(counts, K=3, seed=83, NMF=F)
plot(res5$Ls, ylab="L", main="Smoothed")

res7 <- lda_smoothed(counts, K=3, seed=83, NMF=T)
plot(res6$Ls, ylab="L", main="Smoothed with NMF")
```

```{r}
(Ls <- sapply(list(res1, res2, res3, res4, res5, res6, res7), function(res) res$L))
plot(Ls, ylab="L", main="Compare final L", col=c(1, 1, 2, 1, 2, 3, 3), pch=16)
#black is lda orginal, should be equal
#red is lda with no alpha, should be equal 
#i thought the reshaped versions 4 and 5 would be the same as 2 and 4 respectively, but 1dp off isnt too bad
#interestingly lda_smoothed seems to have performed the best, and nmf initalisation makes it slightly worse?
```

Now we know all 5 work as expected, we can just look at the 3 main ones: lda_orginal_par, lda_reshaped, and lda_smoothed.

# I haven't updated code below this line since editing lda_reshaped and lda_reshaped_noalpha

# Text data
Running these 3 implementations multiple times with different numbers of topics (expecting a peak at K=3), saving the final values of L to `resK`, and keeping the full results of the runs with the highest L.

```{r}
plot_LvK <- function(res, Ks){
  max_line <- data.frame(K=Ks, Max=apply(res, 1, max))
  res <- data.frame(K=Ks, res)
  res_lls <- pivot_longer(res, cols=-"K", names_to="Rep", values_to="L")
  
  p <- ggplot(res_lls, aes(x=K, y=L, group=1))+
    geom_line(data=max_line, aes(x=K, y=Max), colour="grey") +
    geom_point() +
    labs(x="number of topics", y="L") +
    theme_minimal()
  
  return(p)
}
```

```{r, message=FALSE}
Ks <- 2:6
reps <- 2

res1_LvK <- res2_LvK <- res3_LvK <- matrix(NA, length(Ks), reps)
res1_best <- res2_best <- res3_best <- list("L"=-Inf)

for(i in 1:reps){
  for(j in 1:length(Ks)){
    temp <- lda_original_par(docs, K=Ks[j], seed=i)
    res1_LvK[j, i] <- temp$L
    if(temp$L > res1_best$L) res1_best <- temp
    
    temp <- lda_reshaped(counts, K=Ks[j], seed=i)
    res2_LvK[j, i] <- temp$L
    if(temp$L > res2_best$L) res2_best <- temp
    
    temp <- lda_smoothed(counts, K=Ks[j], seed=i)
    res3_LvK[j, i] <- temp$L
    if(temp$L > res3_best$L) res3_best <- temp
  }
}

plot_LvK(res1_LvK, Ks) + labs(title="LDA original")
plot_LvK(res2_LvK, Ks) + labs(title="LDA reshaped")
plot_LvK(res3_LvK, Ks) + labs(title="LDA smoothed")
```

All 3 implementations have a peak at K=3 as expected!

Now comparing the estimated mixing proportions of the best runs

```{r}
plot_mixture <- function(dat, nsamples=10, sample_labels=NULL, topic_label=1:ncol(dat), width=0.9){

  if(is.null(sample_labels)) sample_labels <- paste("doc", 1:nsamples)
  Sample <- factor(sample_labels, levels=rev(unique(sample_labels)))

  plot_dat <- as.data.frame(dat)
  plot_dat <- plot_dat[1:nsamples,]
  colnames(plot_dat) <- paste("Topic", topic_label)
  plot_dat <- cbind(plot_dat, Sample)

  plot_dat <- plot_dat %>%
    pivot_longer(cols=-"Sample", names_to = "Topic", values_to="Proportion")

  p <- ggplot(plot_dat, aes(fill=Topic, x=Sample, y=Proportion)) +
      geom_bar(position="fill", stat="identity", width=width) +
      coord_flip() +
      labs(x="", fill="") +
      theme_minimal() #+
      #theme(text = element_text(size = 14))

  return(p)
}
```

```{r}
plot_mixture(thetas_true) + labs(title="True mixtures")
plot_mixture(res1_best$thetas) + labs(title="LDA original")
plot_mixture(res2_best$thetas) + labs(title="LDA reshaped")
plot_mixture(res3_best$thetas) + labs(title="LDA smoothed")
```

Matching up the topics and comparing the MAE
```{r}
error <- rep(NA, 3)
true_max <- apply(thetas_true, 1, which.max)
mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

for(i in 1:3){
  res <- get(paste0("res", i, "_best"))$thetas
  res <- res / rowSums(res)

  model_max <- apply(res, 1, which.max)
  order <- rep(NA, 3)
  for(j in 1:3){
    samples <- which(true_max == j)
    order[j] <- mode(model_max[samples])
  }
  res <- res[,order]
  error[i] <- mean(abs(thetas_true-res))
}

print(error)
plot(error)
```

So LDA original appears to have performed best here. That could be due to it being able to tune $\alpha$ whereas the other two implementations are given a default value of 1/K (and the true value is 0.1).

Testing tuning alpha
```{r, message=FALSE}
alphas <- c(0.05, 0.1, 1/3, 0.5, 1)
reps <- 2
res_LvA <- matrix(NA, length(alphas), reps)

for(i in 1:reps){
  for(j in 1:length(alphas)){
    temp <- lda_reshaped(counts, K=3, alpha=alphas[j], seed=i*5)
    res_LvA[j, i] <- temp$L
  }
}

plot_LvK(res_LvA, alphas) + labs(title="LDA reshaped", x="alpha")
```

This has a peak at alpha=0.1 as we'd expect!

lda_smoothed also has another hyperparameter $\eta$ which can be tuned, although I dont know what value we'd expect.

# Spectra
Now moving onto the dataset of bacteria spectra

We can no longer use the lda_original methods as these use document vectors, but we can use lda_reshaped which is very similar

For lda_smoothed, compare running with and without NMF initialisation
```{r, message=FALSE}
load("data/Spectra.Rdata")

reps <- 2
K <- 8
res_lls <- matrix(NA, reps*2, 50)
thresh <- 1e-5 #adjusting the threshold because 1e-4 didnt quite seem converged enough

for(i in 1:reps){
  temp <- lda_smoothed(counts, K, seed=i*3, NMF=F, thresh=thresh)
  res_lls[i, 1:length(temp$Ls)] <- temp$Ls
  
  temp <- lda_smoothed(counts, K, seed=i*6, NMF=T, thresh=thresh)
  res_lls[reps+i, 1:length(temp$Ls)] <- temp$Ls
}

colnames(res_lls) <- 1:50

res_lls2 <- res_lls %>%
  as.data.frame %>%
  mutate(Model=factor(c(rep("No NMF", reps), rep("NMF", reps))), 
         Run=factor(rep(1:reps, 2))) %>%
  pivot_longer(cols=-c("Model", "Run"), values_to="L", names_to="Iteration") %>%
  filter(!is.na(L)) %>%
  mutate(Iteration=as.numeric(Iteration))

ggplot(res_lls2, aes(x=Iteration, y=L, color=Model, group=Run)) +
  geom_line() +
  facet_wrap(~Model) +
  theme_minimal()
```
Not sure why NMF makes it perform worse, but we can just continue with the not NMF version.

Try to final optimal K
```{r, message=FALSE}
Ks <- 4:11
reps <- 2

res1_LvK <- res2_LvK <- matrix(NA, length(Ks), reps)
res1_best <- res2_best <- list("L"=-Inf)

for(i in 1:reps){
  for(j in 1:length(Ks)){
    temp <- lda_reshaped(counts, K=Ks[j], seed=i)
    res1_LvK[j, i] <- temp$L
    if(temp$L > res1_best$L) res1_best <- temp
    
    temp <- lda_smoothed(counts, K=Ks[j], seed=i+1)
    res2_LvK[j, i] <- temp$L
    if(temp$L > res2_best$L) res2_best <- temp
  }
}
plot_LvK(res1_LvK, Ks) + labs(title="LDA reshaped")
plot_LvK(res2_LvK, Ks) + labs(title="LDA smoothed")
```

It's good that our results are comparable, but annoying there is not a clear peak. It could be argued that lda_reshaped has an 'elbow' at K=8.

For lda_reshaped try to find optimal alpha
```{r, message=FALSE}
alphas <- c(0.05, 0.1, 1/3, 0.5, 1)
reps <- 2
res_LvA <- matrix(NA, length(alphas), reps)
res3_best <- list("L"=-Inf)

for(i in 1:reps){
  for(j in 1:length(alphas)){
    temp <- lda_reshaped(counts, K=8, alpha=alphas[j], seed=i*j)
    res_LvA[j, i] <- temp$L
    if(temp$L > res3_best$L) res3_best <- temp
  }
}

plot_LvK(res_LvA, alphas) + labs(title="LDA reshaped", x="alpha")
```

Looks like 0.1 is a good value of alpha.

Visualise the results of this model

```{r}
thetas <- res3_best$thetas
#plot_mixture(thetas, nsamples=72, sample_labels = paste(idx, rep(1:8, 9)), width=0.4)
plot_mixture(thetas[seq(1, 72, 8),], nsamples=9, sample_labels=idx[seq(1, 72, 8)], width=0.6)
```

Psm and RO340 are clearly associated with Topics 8 and 1 respectively.

Visualising the results for all the samples using PCA
```{r}
prcomp(thetas)$x[, 1:2] %>%
  as.data.frame %>%
  mutate(Strain=idx) %>%
  ggplot(aes(x=PC1, y=PC2, colour=Strain)) +
    geom_point() +
    theme_minimal() +
    labs(title="PCA of Mixtures")
```

Then compare that to a PCA of the count matrix directly, they are pretty similar...
```{r}
prcomp(counts)$x[, 1:2] %>%
  as.data.frame %>%
  mutate(Strain=idx) %>%
  ggplot(aes(x=PC1, y=PC2, colour=Strain)) +
    geom_point() +
    theme_minimal() +
    labs(title="PCA of Spectra")
```

Plot the topic distributions, and compare to the spectra
```{r}
compare_topic_spectra <- function(beta, counts, mz, topic, strain){
  par(mfrow=c(2, 1))
  plot(mz, beta[topic,], type="l", xlab="m/z", ylab="P(m/z | topic)",
     col="darkgrey", lwd=1.3, main=paste("Topic", topic, "asssociated with", strain))

  matplot(mz, t(counts[idx==strain,]), type="l", ylab="Relative intensity", 
        xlab="m/z", lty=1, main=paste(strain, "spectra"))
}
```

```{r, fig.width=6, fig.height=5}
beta <- res3_best$beta
compare_topic_spectra(beta, counts, mz_locations, topic=8, strain="Psm")

compare_topic_spectra(beta, counts, mz_locations, topic=1, strain="RO340")

compare_topic_spectra(beta, counts, mz_locations, topic=7, strain="RN4220")

compare_topic_spectra(beta, counts, mz_locations, topic=4, strain="Hld")
```


